---
title: "Testing dash on A. White et al bird abundance data"
author: "Kushal K Dey"
date: "9/13/2017"
output: html_document
---

We install and load the **ecostructure** package and load the Himalayan bird abundance data from this package.

```{r warning = FALSE, message = FALSE}
library(ecostructure)
library(Logolas)
library(CountClust)
```

We load the taxonomic data as an ExpressionSet object.

```{r}
data <- get(load(system.file("extdata", "HimalayanBirdsData.rda",
                             package = "ecostructure")))
taxonomic_counts <- t(exprs(data))
rowSums(taxonomic_counts)
m2 <- colMeans(taxonomic_counts)
```

We also load the metadata

```{r}
grid_metadata <- pData(phenoData(data))
head(grid_metadata)
elevation_metadata=grid_metadata$Elevation;
east_west_dir = grid_metadata$WorE;
```

We now apply **dash** on the bird abundance data.

```{r}
system.time(out <- dash(comp_data = taxonomic_counts,
                        optmethod = "mixEM",
                        mode = m2,
                        def_positions = list("center" = Inf, "null" = 1, "corner" = 1),
                        concentration = c(Inf, 100, 50, 20, 10, 5, 2, 1),
                        weight = list("center" = 10, "null" = 1, "corner" = 1),
                        bf = TRUE,
                        verbose=TRUE))
```

We observe how the posterior weight on the singleton relates to the total size of the birds for that forest patch.

```{r}
plot(out$posterior_weights[,1], rowSums(taxonomic_counts), col="red",
     xlab = "Weight of the null component", ylab = "Total number of birds in patch",
     pch = 20, cex = 1)
```

We apply the Grade of Membership (GoM) model on the taxonomic bird abundance data.

### cluster - 2

```{r}
gom_fit_list <- vector(mode = "list", length = 100)
for(m in 1:100){
  gom_fit_2 <- CountClust::FitGoM(taxonomic_counts, K=2, tol=0.1)
  gom_fit_list[[m]] <- gom_fit_2
}
bf_val <- unlist(lapply(gom_fit_list, function(x) return(x$clust_2$BF)))
gom_result <- gom_fit_list[which.min(bf_val)]
```

```{r}
ecostructure::BlockStructure(gom_result[[1]]$clust_2$omega, blocker_metadata = east_west_dir,
                             order_metadata = elevation_metadata,
                             yaxis_label = "Elevation",
                             levels_decreasing = FALSE)

```

```{r}

gom_fit <- CountClust::FitGoM(taxonomic_counts, K=2:4, tol=0.1)

omega <- gom_fit[[1]]$omega
theta <- gom_fit[[1]]$theta

ecostructure::BlockStructure(omega, blocker_metadata = east_west_dir,
                             order_metadata = elevation_metadata,
                             yaxis_label = "Elevation",
                             levels_decreasing = FALSE)

```

### cluster - 3

```{r}

gom_fit <- CountClust::FitGoM(taxonomic_counts, K=2:4, tol=0.1)

omega <- gom_fit[[2]]$omega
theta <- gom_fit[[2]]$theta

ecostructure::BlockStructure(omega, blocker_metadata = east_west_dir,
                             order_metadata = elevation_metadata,
                             yaxis_label = "Elevation",
                             levels_decreasing = FALSE)

```


### cluster - 4

```{r}

gom_fit <- CountClust::FitGoM(taxonomic_counts, K=2:4, tol=0.1)

omega <- gom_fit[[3]]$omega
theta <- gom_fit[[3]]$theta

ecostructure::BlockStructure(omega, blocker_metadata = east_west_dir,
                             order_metadata = elevation_metadata,
                             yaxis_label = "Elevation",
                             levels_decreasing = FALSE)

```


## Refining abundance data after dash adjustment


We now consider those forest patches with small weights on the null component and retain the original counts for those patches. For the ones with large weights on the null component, we use a shrunk estimate of the counts of bird species.

```{r}
idx <- which(out$posterior_weights[,1] < 0.01)
tmp <- matrix(0, dim(out$datamean)[1], dim(out$datamean)[2])
tmp[idx, ] <- out$datamean[idx, ]
tmp[-idx, ] <- out$posmean[-idx, ]

mod_counts <- round((rowSums(taxonomic_counts) %*% t(rep(1, dim(taxonomic_counts)[2])))*tmp)
rownames(mod_counts) <- rownames(taxonomic_counts)
colnames(mod_counts) <- colnames(taxonomic_counts)
```


The Grade of Membership model fit on this refined counts matrix is as follows

### cluster - 2

```{r}

gom_fit_2 <- CountClust::FitGoM(mod_counts, K=2:4, tol=0.1)
omega2 <- gom_fit_2[[1]]$omega
theta2 <- gom_fit_2[[1]]$theta

BlockStructure(omega2, blocker_metadata = east_west_dir,
               order_metadata = elevation_metadata,
               yaxis_label = "Elevation",
               levels_decreasing = FALSE)
```

### cluster - 3

```{r}

gom_fit_2 <- CountClust::FitGoM(mod_counts, K=2:4, tol=0.1)
omega2 <- gom_fit_2[[2]]$omega
theta2 <- gom_fit_2[[2]]$theta

BlockStructure(omega2, blocker_metadata = east_west_dir,
               order_metadata = elevation_metadata,
               yaxis_label = "Elevation",
               levels_decreasing = FALSE)
```


### cluster - 4

```{r}

gom_fit_2 <- CountClust::FitGoM(mod_counts, K=2:4, tol=0.1)
omega2 <- gom_fit_2[[3]]$omega
theta2 <- gom_fit_2[[3]]$theta

BlockStructure(omega2, blocker_metadata = east_west_dir,
               order_metadata = elevation_metadata,
               yaxis_label = "Elevation",
               levels_decreasing = FALSE)
```

```{r}
gom_fit_list <- vector(mode = "list", length = 100)
for(m in 1:100){
  gom_fit_2 <- CountClust::FitGoM(mod_counts, K=2, tol=0.1)
  gom_fit_list[[m]] <- gom_fit_2
}
bf_val <- unlist(lapply(gom_fit_list, function(x) return(x$clust_2$BF)))
gom_result <- gom_fit_list[which.min(bf_val)]
```
